---
title: "How does fuseMLR work?"
author: "Cesaire Fouodo"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{How does fuseMLR work?}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## A - Introduction #

The R package **fuseMLR** offers a framework for late integrative predictive modeling using multi-omics data. 
This vignette serves as a user guide to use the package effectively.
Three key functionalities are provided: variable selection for modality-specific data, late integrative multi-omics training, and predictions for testing multi-omics datasets. 

## B - Data

```{r libraries, warning = FALSE}
library(fuseMLR)
```

The following example is based on simulated data available in `fuseMLR`. Data have been simulated using the R package `InterSIM`, version 2.2.0. Two types of data were simulated: training and testing datasets. Each consists of four `data.frame`s—gene expression, protein expression, methylation data, and target variable observations. Individuals are organized in rows, variables in columns, with an additional column for individual IDs. In total, $70$ individuals with $50$ individuals pro layer have been simulated for training, and $23$ ($20$ per layer) for testing. Individuals do not necessarily overlapped. Effects have been introduced for gene expression and methylation by shifting the means by $0.5$ to create case-control study with $50$% prevalence. Individuals do not necessarily overlap. Effects were introduced in gene expression and methylation by shifting the means by 0.5 to create a case-control study. For illustration, the number of variables was kept smaller than what is typically expected in reality. The data simulation code is available [here](https://github.com/imbs-hl/fuseMLR/blob/master/test_code/build_data.R).

```{r data_exam, include=TRUE, eval=TRUE}
data("entities")
# This is a list containing two lists of data: training and test.
# Each sublist contains three entities.
str(object = entities, max.level = 2L)
```

## C - Training #

The training process is handled in **fuseMLR** by a class called **Training**. We will use ```training``` to refer to an instance from this class. Function ```createTraining``` is available to create an empty (without layers) ```training```.

```{r training, include=TRUE, eval=TRUE}
training <- createTraining(id = "training",
                           ind_col = "IDS",
                           target = "disease",
                           target_df = entities$training$target,
                           verbose = FALSE)
print(training)
```

A ```training``` contains modality-specific training layers (instances of the **TrainLayer** class) and a meta-layer (instance of the **TrainMetaLayer** class). The modality-specific training layers encapsulate training data modalities, variable selection functions and learners (i.e. ```R``` functions implementing statistical predicting methods). The meta-layer encapsulates the meta-learner.

## C.1 - Creating a training ##

We will use ```training``` (resp. ```testing```) to refer to an instance from the **Training** (resp. **Testing**) class. A ```training``` contains modality-specific training layers (instances of the **TrainLayer** class) and a meta-layer (instance of the **TrainMetaLayer** class). The modality-specific training layers encapsulate training data modalities, variable selection functions and learners (i.e. ```R``` functions implementing statistical predicting methods). The meta-layer encapsulates the meta-learner.

Three main functions are necessary to create ```training```: ```createTraining()```, ```createTrainLayer()``` and ```createTrainMetaLayer()```. ```createTraining()``` creates an empty ```training```, on which modalitity-specific training layers can be add using the function ```createTrainLayer()```. Function ```createTrainMetaLayer()``` is used to add meta-layer to ```training```. 

The following code adds a gene expression, a protein abundance, and a methylation layer to ```training```.

```{r training_layers, include=TRUE, eval=TRUE}
# Create gene expression layer
createTrainLayer(training = training,
                 train_layer_id = "geneexpr",
                 train_data = entities$training$geneexpr,
                 varsel_package = "Boruta",
                 varsel_fct = "Boruta",
                 varsel_param = list(num.trees = 1000L,
                                     mtry = 3L,
                                     probability = TRUE),
                 lrner_package = "ranger",
                 lrn_fct = "ranger",
                 param_train_list = list(probability = TRUE,
                                         mtry = 1L),
                 param_pred_list = list(),
                 na_rm = TRUE)

# Create gene protein abundance layer
createTrainLayer(training = training,
                 train_layer_id = "proteinexpr",
                 train_data = entities$training$proteinexpr,
                 varsel_package = "Boruta",
                 varsel_fct = "Boruta",
                 varsel_param = list(num.trees = 1000L,
                                     mtry = 3L,
                                     probability = TRUE),
                 lrner_package = "ranger",
                 lrn_fct = "ranger",
                 param_train_list = list(probability = TRUE,
                                         mtry = 1L),
                 param_pred_list = list(type = "response"),
                 na_rm = TRUE)

# Create methylation layer
createTrainLayer(training = training,
                 train_layer_id = "methylation",
                 train_data = entities$training$proteinexpr,
                 varsel_package = "Boruta",
                 varsel_fct = "Boruta",
                 varsel_param = list(num.trees = 1000L,
                                     mtry = 3L,
                                     probability = TRUE),
                 lrner_package = "ranger",
                 lrn_fct = "ranger",
                 param_train_list = list(probability = TRUE,
                                         mtry = 1L),
                 param_pred_list = list(),
                 na_rm = TRUE)
```

Also add a meta layer.

```{r training_meta_layers, include=TRUE, eval=TRUE}
# Create meta layer
createTrainMetaLayer(training = training,
                     meta_layer_id = "meta_layer",
                     lrner_package = NULL,
                     lrn_fct = "weightedMeanLearner",
                     param_train_list = list(),
                     param_pred_list = list(),
                     na_rm = FALSE)
```

Function ```upsetplot()``` is available to generate an upset of the training data, i.e. an overview how patients overlap across layers.

```{r upsetplot, include=TRUE, eval=TRUE, }
upsetplot(object = training, order.by = "freq")
```

## C.2 - Variable selection ##

Function ```varSelection()``` performs modality-specific variable selection.

```{r varsel, include=TRUE, eval=TRUE}
# Variable selection
set.seed(5467)
var_sel_res <- varSelection(training = training,
                            verbose = FALSE)
print(var_sel_res)
```

For each layer, the variable selection results show the chosen variables. In this example, we perform variable selection. Users can opt to conduct variable selection on individual layers if desired.

## C.2 - Training ##

We use the function ```fusemlr()``` to train our models using the subset of selected variables.

```{r lrner_train, include=TRUE, eval=TRUE}
set.seed(5462)
fusemlr(training = training,
        use_var_sel = TRUE,
        verbose = FALSE)

print(training)
# See also summary(training)
```

We use `extractModel()` to retrieve the list of stored models and `extractData()` to retrieve training data.

```{r basic_lrnr, include=TRUE, eval=TRUE}
models_list <- extractModel(training = training)
data_list <- extractData(training = training)
```


# D - Predicting #

In this section, we create a ```testing``` instance (from the *Testing* class) and make predictions for new data. This is done analogously to ```training```. The only difference that only the testing data modalities are required. Relevant functions are ```createTesting()``` and ```createTestLayer()```.

```{r testing, include=TRUE, eval=TRUE}
# Create testing for predcitions
testing <- createTesting(id = "testing",
                         ind_col = "IDS")

# Create gene expression layer
createTestLayer(testing = testing,
                test_layer_id = "geneexpr",
                test_data = entities$testing$geneexpr)

# Create gene protein abundance layer
createTestLayer(testing = testing,
                test_layer_id = "proteinexpr",
                test_data = entities$testing$proteinexpr)

# Create methylation layer
createTestLayer(testing = testing,
                test_layer_id = "methylation",
                test_data = entities$testing$proteinexpr)
```

We can also generate an upset plot to visualize patient overlap across testing layers.

```{r upsetplot_new, include=TRUE, eval=TRUE, }
upsetplot(object = testing, order.by = "freq")
```

Function ```predict()``` is available for predicting.

```{r new_pred, include=TRUE, eval=TRUE}
predictions <- predict(object = training, testing = testing)
print(predictions)
```

- Prediction performances for layer-specific available patients, and all patients on the meta layer.

```{r performance_all, include=TRUE, eval=TRUE}
pred_values <- predictions$predicted_values
actual_pred <- merge(x = pred_values,
                     y = entities$testing$target,
                     by = "IDS",
                     all.y = TRUE)
x <- as.integer(actual_pred$disease == 2L)

# On all patients
perf_estimated <- sapply(X = actual_pred[ , 2L:5L], FUN = function (my_pred) {
  bs <- bs <- mean((x[complete.cases(my_pred)] - my_pred[complete.cases(my_pred)])^2)
  return(bs)
})
print(perf_estimated)
```

- Prediction performances for overlapping individuals.

```{r performance_overlap, include=TRUE, eval=TRUE}
# On overlapping patients
perf_overlapping <- sapply(X = actual_pred[complete.cases(actual_pred),
                                           2L:5L],
                           FUN = function (my_pred) {
                             bs <- mean((x[complete.cases(actual_pred)] - my_pred)^2)
                             return(bs)
                           })
print(perf_overlapping)
```

Note that our example is based on simulated data for usage illustration; only one run is not enough to appreciate the performances of our models.

# E - Interface and wrapping #

We distinguish common supervised learning arguments from method specific arguments. The common arguments are a matrix ```x``` of independent variables and  ```y``` representing a response variable. These arguments are handled by ```fuseMLR```, so users do not need to make any adjustments themselves. We also define standard argument names for predicting. The arguments ```object``` and ```data``` (used by the generic ```R``` function ```predict``` to pass model and the data for which prediction are needed) are also stanard in ```fuseMLR```. If the original learning or predicting function do not use these names as arguments, either an interface or a wrapping of the original function can be done to solve name discepancies. 

## Interface ##

The interface approach leverages the arguments in ```createTrainLayer()``` to map the argument names of the original learning function. In the example below, the gene expression layer is re-created using the ```svm``` (Support Vector Machine) function from the ```e1071``` package as the learner. A discrepancy arises in the argument names of the ```predict.svm``` function, which uses ```object``` and ```newdata```.

```{r interface, include=TRUE, eval=TRUE}
# Remove the current gene expression layer from training
removeLayer(training = training, layer_id = "geneexpr")
# Re-create the gene expression layer with support vector machine as learner.
createTrainLayer(training = training,
                 train_layer_id = "geneexpr",
                 train_data = entities$training$geneexpr,
                 varsel_package = "Boruta",
                 varsel_fct = "Boruta",
                 varsel_param = list(num.trees = 1000L,
                                     mtry = 3L,
                                     probability = TRUE),
                 lrner_package = "e1071",
                 lrn_fct = "svm",
                 param_train_list = list(type = 'C-classification',
                                         kernel = 'radial',
                                         probability = TRUE),
                 param_pred_list = list(probability = TRUE),
                 na_rm = TRUE,
                 x = "x",
                 y = "y",
                 object = "object",
                 data = "newdata", # Name discrepancy resolved.
                 extract_pred_fct = function (pred) { 
                   pred <- attr(pred, "probabilities")
                   return(pred[ , 1L])
                 }
)
# Variable selection
set.seed(5467)
var_sel_res <- varSelection(training = training,
                            verbose = FALSE)
set.seed(5462)
training <- fusemlr(training = training,
                    use_var_sel = TRUE,
                    verbose = FALSE)

print(training)
```

## Wrapping ##

In the wrapping approach we re-define the function ```mylasso``` to run a Lasso regression from the ```glmnet``` package as the meta-leaner.

```{r wrap, include=TRUE, eval=FALSE}
# We wrap the original functions
mylasso <- function (x, y,
                     nlambda = 25,
                     nfolds = 5) {
  # Perform cross-validation to find the optimal lambda
  cv_lasso <- cv.glmnet(x = as.matrix(x), y = y,
                        family = "binomial",
                        type.measure = "deviance",
                        nfolds = nfolds)
  best_lambda <- cv_lasso$lambda.min
  lasso_best <- glmnet(x = as.matrix(x), y = y,
                       family = "binomial",
                       alpha = 1,
                       lambda = best_lambda
  )
  lasso_model <- list(model = lasso_best)
  class(lasso_model) <- "mylasso"
  return(lasso_model)
}

predict.mylasso <- function(object, data) {
  glmnet_pred <- predict(object = object$model,
                         newx = as.matrix(data),
                         type = "response",
                         s = object$model$lambda)
  return(as.vector(glmnet_pred))
}

# Remove the current gene expression layer from training
removeLayer(training = training, layer_id = "meta_layer")
# Re-create the gene expression layer with support vector machine as learner.
createTrainMetaLayer(training = training,
                     meta_layer_id = "meta_layer",
                     lrner_package = NULL,
                     lrn_fct = "mylasso",
                     param_train_list = list(nlambda = 100L),
                     na_rm = TRUE)
set.seed(5462)
training <- fusemlr(training = training,
                    use_var_sel = TRUE,
                    verbose = FALSE)
print(training)
```

### Appendix

In addition to any pre-existing learner in R as a meta-learner, we have implemented the following ones.

```{r implemented_learners, include=TRUE, echo=FALSE}
# Load knitr package
library(knitr)

# Create a data frame
data <- data.frame(
  Leaner = c("weightedMeanLearner", "bestSpecificLearner"),
  Description = c("The weighted mean meta learner. It uses meta data to estimate the weights of the modality-specific models", "The best layer-specific model is used as meta model.")
)

# Generate the table
kable(data, caption = "")
```

&copy; 2024 Institute of Medical Biometry and Statistics (IMBS). All rights reserved.
